---
title: Supplementary Material
editor: visual
---

```{r}
#| label: setup-supplement
#| include: false
knitr::opts_chunk$set(dev = 'png', echo = FALSE, results = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(sf)
library(stars)
library(ggridges)
library(patchwork)
library(units)
library(rnaturalearth)
library(rnaturalearthdata)
library(here)
library(eks)

theme_set(theme_bw())

# define the study area
asia_bbox_sf <- st_bbox(c(xmin = 60, xmax = 150, ymin = -11, ymax = 54), crs = 4326)
world_bbox_sf <- st_bbox(c(xmin = -130, xmax = 155, ymin = -40, ymax = 60), crs = 4326)

# get country boundaries shapefile for plotting
sf_use_s2(FALSE)
countries <- ne_countries(scale = 'large') %>% 
  st_crop(asia_bbox_sf)
countries_small<- ne_countries() %>%
  st_crop(asia_bbox_sf)
coasts <- ne_coastline() %>%
  st_crop(asia_bbox_sf)
coasts_world <- ne_coastline() %>% 
  st_crop(world_bbox_sf)
countries_world <- ne_countries(scale = 'large') %>% 
  st_crop(world_bbox_sf)
sf_use_s2(TRUE)

country_stats <- readRDS('data/data-derived/country_stats.rds') # country level stats

load(here('data/data-derived/occurrence_data.RData'))

arch_trace_dat <- readRDS(here('data/data-derived/arch_trace.rds')) |>
  mutate(mil = (med_age - med_age %% 1000) / 1000 + 1) |>
  filter(between(mil, -9, -2)) |> # reduces to 542 unique sites
  mutate(mil = as.factor(mil)) 

fig2_dat <- irri_occ %>%
  st_drop_geometry() %>%
  select(bio1, bio5, bio10) %>%
  mutate(across(everything(), ~units::set_units(.x, 'degree_C')))

quantiles <- c(0.01, 0.025, 0.05, 0.1, 0.9, 0.95, 0.975, 0.99)

quants_mod <- fig2_dat %>%
  reframe(across(everything(), 
                   ~quantile(.x, quantiles, type = 8))) %>%
  round(1) %>%
  mutate(quantile = quantiles, .before = 1)

quant_975 <- filter(quants_mod, quantile == 0.975)
bio1_limit <- quant_975$bio1
bio5_limit <- quant_975$bio5
bio10_limit <- quant_975$bio10

cmip6_supp <- readRDS(here('data/data-derived/cmip6_data.rds'))
```

# Supplementary Materials

Supplementary Materials for **Projected warming will exceed the long-term thermal limits of rice cultivation**

Nicolas Gauthier, Ornob Alam, Michael D. Purugganan, Jade d'Alpoim Guedes

Corresponding author: nicolas.gauthier\@ufl.edu

**This PDF file includes:**

-   Figs. S1 to S8

-   Tables S1 and S2

**Other Supplementary Materials for this manuscript include the following:**

-   Data S1 to S3

## Fig. S1.

```{r}
#| label: fig-current-limits
#| fig-width: 6
#| fig-height: 5
#| cache: true
#| fig-unnumbered: true
#| fig-cap: "Present day geographic distribution of temperature thresholds in CHELSA V2.1 data. A) Mean annual temperature greater than 28.2째C. B) Mean maximum temperature of the warmest month greater than 40.4째C. C) Mean maximum temperature of the warmest quarter greater than 32.7째C."
chelsa2 <- c('CHELSA_bio1_1981-2010_V.2.1.tif', 
            'CHELSA_bio5_1981-2010_V.2.1.tif', 
            'CHELSA_bio10_1981-2010_V.2.1.tif') %>%
  here('data/data-raw', .) |>
  read_stars(proxy = TRUE) |>
  st_crop(st_bbox(coasts_world))

current_limits_1 <- ggplot() +
  geom_stars(data = chelsa2[1] >= units::drop_units(bio1_limit), downsample =  5) +
  scale_fill_manual(values = c('#fee8c8', '#e34a33'), 
                    name = paste0('MAT > ', bio1_limit), 
                    na.value = NA, na.translate = FALSE) +
  geom_sf(data = coasts_world) +
  labs(x = '', y = '') +
  coord_sf(expand = FALSE)

current_limits_5 <- ggplot() +
  geom_stars(data = chelsa2[2] >= units::drop_units(bio5_limit), downsample =  5) +
  scale_fill_manual(values = c('#fee8c8', '#e34a33'), 
                    name = paste0('TMAX > ', bio5_limit), 
                    na.value = NA, na.translate = FALSE) +
  geom_sf(data = coasts_world) +
  labs(x = '', y = '') +
  coord_sf(expand = FALSE)

current_limits_10 <- ggplot() +
  geom_stars(data = chelsa2[3] >= units::drop_units(bio10_limit), downsample =  5) +
    scale_fill_manual(values = c('#fee8c8', '#e34a33'), 
                       name = paste0('TWARM > ', bio10_limit),
                    na.value = NA, na.translate = FALSE) +
  geom_sf(data = coasts_world) +
  labs(x = '', y = '') +
  coord_sf(expand = FALSE)

current_limits_1 / current_limits_5 / current_limits_10 + plot_annotation(tag_levels = 'A', tag_suffix = ')') 
```

{{< pagebreak >}}

## Fig. S2.

```{r}
#| label: fig-bootstrap
#| fig-width: 6
#| fig-height: 4
#| fig-unnumbered: true
#| fig-cap: "A) Bootstrapped kernel density estimates of rice's thermal extent throughout the Holocene. Bootstrap resampling accounts for uncertainty in radiocarbon and phase-based dating of archaeological rice finds. Contemporary temperature thresholds indicated in red. B) Estimated 97.5% quantiles from the bootstrapped dataset. Contemporary quantiles indicated in red."
trace_boot <- readRDS('data/data-derived/trace_boot.rds')

# contemporary quantiles for plotting
quant_dat <- quant_975 |>
               select(-quantile) |>
               setNames(c('MAT', 'tmax', 'twarm')) |>
              pivot_longer(everything())

boot_1 <- trace_boot %>%
  filter(time <= -2000) %>%
  select(rep, MAT:twarm) %>%
  pivot_longer(MAT:twarm) %>%
  group_by(rep, name) %>%
  tidy_kde() %>%
ggplot(aes(value, estimate)) +
  geom_line(alpha = 0.01, aes(group = rep)) +
  facet_wrap(~name, scales = 'free_y', nrow = 3) +
  geom_vline(data = quant_dat, aes(xintercept = as.numeric(value)), linetype = 2, color = 'red') +
  labs(x = 'Temperature')

boot_2 <- trace_boot %>%
  filter(time <= -2000) %>%
  pivot_longer(MAT:twarm, values_to = 'temperature') %>%
  mutate(temperature = units::set_units(temperature, 'degree_c')) %>%
  group_by(rep, name) %>%
  reframe(value = quantile(temperature, 0.975, type = 8)) %>%
  ggplot(aes(value)) +
    geom_histogram() +
    facet_wrap(~name, scales = 'free_x', nrow = 3) +
    labs(x = 'Temperature') +
    theme_bw() +
    geom_vline(data = quant_dat, aes(xintercept = value), linetype = 2, color = 'red')

rm(trace_boot, quant_dat)

boot_1 + boot_2 +  plot_annotation(tag_levels = 'A')
```

{{< pagebreak >}}

## Fig. S3.

```{r}
#| label: fig-recon-timeseries
#| fig-width: 12
#| fig-height: 6
#| fig-unnumbered: true
#| fig-cap: "Comparison of Holocene temperature trends over Asia from transient climate models and reconstructions. A) Holocene trends in mean annual temperature, relative to the 3-5ka mean, from Osman et al. 2021 (black, one standard deviation uncertainty in grey), Erb et al. 2022 (blue), and CHELSA-TraCE21k (red). B) Comparison of annual, seasonal, and monthly temperature trends over Asia from CHELSA-TraCE21k."
trace_ts <- read_rds(here('data/data-derived/trace_ts.rds'))

a <- trace_ts |>
  pivot_longer(-time) |>
  mutate(name = factor(case_when(name == 'bio1' ~ 'Mean annual temperature',
                                 name == 'bio5' ~  'Max temperature of warmest month',
                                 name == 'bio10' ~  'Max temperature of the warmest quarter'),
                        levels = c('Mean annual temperature', 
                                   'Max temperature of the warmest quarter',
                                   'Max temperature of warmest month'))) |>
  ggplot() +
  geom_line(aes(time, value, group = name)) +
  facet_wrap(~name, scales = 'free_y', ncol = 1) +
  labs(x = 'Time (ka BP)', y = 'Temperature (C)') +
  scale_x_reverse(breaks = seq(0, 12, 2)) +
  theme_bw()

# Calculate the area-weighted averaged temperatures and uncertainties from the Erb et al. reconstruction and save the results.
erb_rast <- read_rds(here('data/data-derived/erb_rast.rds'))

erb_area_total <- st_area(erb_rast) |>
  pull(area) |> 
  sum()

erb_ts <- (erb_rast * st_area(erb_rast)) |>
  st_apply(3, sum)  |>
  as_tibble() |>
  mutate(across(-ages, ~units::drop_units(.x / erb_area_total)),
         time = ages / 1000) |>
  select(-ages)

# Do the same for the LGMR, although this is more complicated as we have to calculate the anomalies by hand for comparison to Erb.

lgmr_rast <- read_rds('data/data-derived/lgmr_rast.rds')

lgmr_mean_reference <- lgmr_rast |>
  filter(between(time, 3000, 5000)) |>
  st_apply(1:2, mean)

lgmr_sd_reference <- lgmr_rast |>
  filter(between(time, 3000, 5000)) |> 
  select(sat_std) |>
  # we need to account for uncertainty in the reference period too
  st_apply(1:2, function(x) sqrt(sum(x^2) / 11) / sqrt(11)) # 11 is the number of time steps

lgmr_uncertainty <- sqrt(lgmr_rast['sat_std'] ^ 2 + lgmr_sd_reference ^ 2)

lgmr_rast_anom <- lgmr_rast['sat'] - lgmr_mean_reference['sat']

lgmr_anomalies <- c(anom = lgmr_rast_anom,
  anom_upper = lgmr_rast_anom + lgmr_uncertainty,
  anom_lower = lgmr_rast_anom - lgmr_uncertainty)

lgmr_area_total <- st_area(lgmr_rast_anom) |>
  as_tibble() |>
  pull(area) |>
  sum()

lgmr_anom_ts <- (lgmr_anomalies * st_area(lgmr_rast_anom)) |>
  st_apply(3, sum)  |>
  as_tibble() |>
    mutate(across(-time, ~units::drop_units(.x / lgmr_area_total)),
           time = time / 1000)

b <- ggplot(data = lgmr_anom_ts, aes(x = time)) +
  geom_line(aes(y = anom)) +
  geom_ribbon(aes(ymin = anom_lower, ymax = anom_upper), alpha = 0.1) +
  geom_line(data = trace_ts, aes(x = time, y = bio1 - 19.737), color = 'red') +
    geom_line(data = erb_ts, aes(y = sat_mean), color = 'blue') +
    geom_ribbon(data = erb_ts, aes(ymin = sat_low , ymax = sat_high), alpha = 0.1, fill = 'blue') +
  scale_x_reverse(breaks = seq(0, 12, 2))  +
  labs(x = 'Time (ka BP)', y = 'Temperature anomaly (C)') 

b + a + plot_layout(guides = 'collect', axes = 'collect') + plot_annotation(tag_levels = 'A', tag_suffix = ')')
```

{{< pagebreak >}}

## Fig. S4.

```{r}
#| label: fig-kde-lgmr
#| fig-width: 10
#| fig-height: 6
#| fig-unnumbered: true
#| fig-cap: "Kernel density estimates of the thermal distribution by millennium across two different temperature datasets, with raw data points indicated in black and contemporary rice thermal limits in red. A) Temperature data are derived from CHELSA-TraCE21k V1.0 (Karger et al. 2023) (same as Figure 3b in main text). B) Temperature data derived from the Last Glacial Maximum Reanalysis (LGMR) dataset (Osman et al. 2022). C) Difference in kernel density estimates of mean annual temperature (MAT) between the CHELSA-TraCE21k and LGMR datasets."
new_lgmr_times <- lgmr_rast |>
  st_get_dimension_values('time') |>
  years() %>%
  `-`(ym('1950-01'), .)

lgmr2 <- lgmr_rast |>
  st_set_dimensions('time', values = (new_lgmr_times)) %>%
  .[,,,61:1]

lgmr_extract_test <- arch_trace_dat |>
  mutate(time = ym('1950-01') - years(as.integer(med_age) * -1)) %>%
  st_extract(lgmr2, ., time_column = 'time', interpolate_time = TRUE) %>%
  st_drop_geometry() %>%
  bind_cols(arch_trace_dat, .) |>
  remove_missing()

arch_dens_mat <- arch_trace_dat %>%
  ggplot(aes(MAT, mil)) +
  scale_y_discrete(limits = rev) +
  geom_density_ridges(aes(fill = mil), 
                      panel_scaling = TRUE, 
                      scale = 2, 
                      alpha = 0.5, 
                      jittered_points = TRUE, 
                      point_alpha = 0.25, 
                      point_size = 0.5) +
  labs(y = 'Millennium BP', x = 'Mean annual temperature') +
  geom_vline(xintercept = as.numeric(bio1_limit), linetype = 2, color = 'red') +
  scale_fill_brewer(palette = 'Spectral', guide = 'none') +
  scale_x_continuous(limits = c(0, 32))

arch_dens_lgmr <- lgmr_extract_test %>%
  ggplot(aes(sat, mil)) +
  scale_y_discrete(limits = rev) +
  geom_density_ridges(aes(fill = mil), 
                      panel_scaling = TRUE, 
                      scale = 2, 
                      alpha = 0.5, 
                      jittered_points = TRUE, 
                      point_alpha = 0.25, 
                      point_size = 0.5) +
  labs(y = 'Millennium BP', x = 'Mean annual temperature') +
  geom_vline(xintercept = as.numeric(bio1_limit), linetype = 2, color = 'red') +
  scale_fill_brewer(palette = 'Spectral', guide = 'none') +
  scale_x_continuous(limits = c(0, 32))

# the difference between Chelsa-TraCE and LGMR ensemble mean at each point in time/space
arch_dens_diff <- lgmr_extract_test %>%
  ggplot(aes(sat - MAT, mil)) +
  scale_y_discrete(limits = rev) +
  geom_density_ridges(aes(fill = mil), 
                      panel_scaling = TRUE, 
                      scale = 2, 
                      alpha = 0.5, 
                      jittered_points = TRUE, 
                      point_alpha = 0.25, 
                      point_size = 0.5) +
  labs(y = 'Millennium BP', x = 'Difference (C)') +
  geom_vline(xintercept = 0, linetype = 2, color = 'red') +
  scale_fill_brewer(palette = 'Spectral', guide = 'none')

# why don't the tags work here?
(arch_dens_mat / arch_dens_lgmr + plot_layout(axes = 'collect')) | arch_dens_diff + plot_layout(widths = c(1, .1))  + plot_annotation(tag_levels = 'A', tag_suffix = ')') 
```

{{< pagebreak >}}

## Fig. S5.

```{r}
#| label: fig-country-summaries
#| fig-width: 8
#| fig-height: 6
#| fig-unnumbered: true
#| fig-cap: "Projected increases in land area surpassing key temperature thresholds in each of the top 15 rice-producing countries in east Asia."
country_stats %>% 
  ggplot(aes(ssp, percent, group = variable, color = variable)) +
  facet_wrap(~name) +
  scale_color_manual(values = c('#abdda4', '#fdae61', '#d7191c'
)) +
  geom_line(linewidth = .8) +
  geom_point() +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
```

{{< pagebreak >}}

## Fig. S6.

```{r}
#| label: fig-niche-all-models-bio1
#| cache: true
#| fig-width: 6
#| fig-height: 5
#| fig-unnumbered: true
#| fig-cap: "Predicted changes in mean annual temperature over currently cultivated rice areas in Asia for multiple CMIP6 ensemble members."
ggplot() +
  geom_stars(data = slice(cmip6_supp, 'variable', 1) > units::drop_units(bio1_limit)) +
  facet_grid(model~ssp) +
  scale_fill_manual(values = c('#fee8c8', '#e34a33'), 
                    name = paste0('MAT > ', bio1_limit), 
                    na.value = NA, na.translate = FALSE) +
  geom_sf(data = coasts) +
  labs(x = '', y = '') +
  theme(
    axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 5)
  )
```

{{< pagebreak >}}

## Fig. S7.

```{r}
#| label: fig-niche-all-models-bio10
#| cache: true
#| fig-width: 6
#| fig-height: 5
#| fig-unnumbered: true
#| fig-cap: "Predicted changes in mean maximum temperature of the warmest quarter over currently cultivated rice areas in Asia for multiple CMIP6 ensemble members."
ggplot() +
  geom_stars(data = slice(cmip6_supp, 'variable', 2) > units::drop_units(bio10_limit)) +
  facet_grid(model~ssp) +
  scale_fill_manual(values = c('#fee8c8', '#e34a33'), 
                       name = paste0('TWARM > ', bio10_limit),
                    na.value = NA, na.translate = FALSE) +
  geom_sf(data = coasts) +
  labs(x = '', y = '')+
  theme(
    axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 5)
  )
```

{{< pagebreak >}}

## Fig. S8.

```{r}
#| label: fig-niche-all-models-bio5
#| cache: true
#| fig-width: 6
#| fig-height: 5
#| fig-unnumbered: true
#| fig-cap: "Predicted changes in mean maximum temperature of the warmest month over currently cultivated rice areas in Asia for multiple CMIP6 ensemble members."
ggplot() +
  geom_stars(data = slice(cmip6_supp, 'variable', 3) > units::drop_units(bio5_limit)) +
  facet_grid(model~ssp) +
  scale_fill_manual(values = c('#fee8c8', '#e34a33'), 
                    name = paste0('TMAX > ', bio5_limit), 
                    na.value = NA, na.translate = FALSE) +
  geom_sf(data = coasts) +
  labs(x = '', y = '') +
  theme(
    axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 5)
  )
```

{{< pagebreak >}}

## Fig. S9.

```{r}
#| label: fig-displacement
#| fig-width: 6
#| fig-height: 3
#| fig-unnumbered: true
#| fig-cap: "Estimated displacement of the optimal growing location of rice landraces away from the equator, based on genetic offset analysis across three future temperature scenarios. The x-axis indicates the difference between the absolute projected latitude in the future scenario and the absolute latitude today, with positive values indicating a shift in the optimal growing location of a given population away from the equator."
list(japonica = readRDS(here('data/data-derived/japonica_lat_shift.rds')),
     indica = readRDS(here('data/data-derived/indica_lat_shift.rds'))) |>
  bind_rows(.id = 'landrace') |>
  mutate(ssp = case_when(name == 'diff2.6' ~ 'SSP1-2.6',
                         name == 'diff7.0' ~ 'SSP3-7.0',
                         name == 'diff8.5' ~ 'SSP5-8.5')) |>
ggplot(aes(x = value, fill = ssp)) +
  geom_density(alpha = 0.3) +
  facet_grid(landrace ~ ssp) +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_fill_brewer(palette = 'Spectral', direction = -1) +
  labs(x = 'Latitudinal displacement (degrees)')
```

{{< pagebreak >}}

## Table S1.

```{r}
#| label: tbl-quantiles
#| tbl-cap: "Present-day rice extent quantiles (in 째C) across three datasets of varying resolution and spatiotemporal support, compared to those calculated from archaeological sites in Asia throughout the Holocene."
#| tbl-unnumbered: true
quants_mod_monfreda <- monfreda_occ %>%
  st_drop_geometry() %>%
  select(bio1, bio5, bio10) %>%
  mutate(across(everything(), ~set_units(.x, 'degree_C'))) %>%
  reframe(across(everything(), 
                   ~quantile(.x, quantiles, type = 8))) %>%
  round(1) %>%
  mutate(quantile = quantiles, .before = 1)

quants_mod_gbif <- occ_dat %>%
  st_drop_geometry() %>%
  select(bio1, bio5, bio10) %>%
  mutate(across(everything(), ~set_units(.x, 'degree_C'))) %>%
  reframe(across(everything(), 
                   ~quantile(.x, quantiles, type = 8))) %>%
  round(1) %>%
  mutate(quantile = quantiles, .before = 1)

quants_arch <- arch_trace_dat %>%
  st_drop_geometry() %>%
  select(bio1 = MAT, bio5 = tmax, bio10 = twarm) %>%
  mutate(across(everything(), ~set_units(.x, 'degree_C'))) %>%
  reframe(across(everything(), 
                   ~quantile(.x, quantiles, type = 8))) %>%
  round(1) %>%
  mutate(quantile = quantiles, .before = 1)


left_join(quants_mod, quants_mod_monfreda, by = 'quantile') %>%
 left_join(quants_mod_gbif, by = 'quantile') %>%
 left_join(quants_arch, by = 'quantile') %>%
 drop_units() %>% # do we want celsius units?
 mutate(quantile = set_units(quantile * 100, '%')) %>%
 select(quantile, bio1.x, bio1.y, bio1.x.x, bio1.y.y, bio5.x, bio5.y, bio5.x.x, bio5.y.y, bio10.x, bio10.y, bio10.x.x, bio10.y.y) %>%
  setNames(c('Quantile', rep(c('IRRI', 'Mon.', 'GBIF', 'Arch.'), 3))) %>%
  knitr::kable(format = 'html') |>
  #knitr::kable(format = fmt, booktabs = knitr::is_latex_output(), align = "c") |>
  kableExtra::add_header_above(tibble(names = c('', 'MAT', 'TMAX', 'TWARM'), spans = c(1,4,4,4)))
 # kableExtra::kable_styling(full_width = FALSE, latex_options = c("hold_position"))

# df_quant <- left_join(quants_mod, quants_mod_monfreda, by = "quantile") %>%
#   left_join(quants_mod_gbif, by = "quantile") %>%
#   left_join(quants_arch,       by = "quantile") %>%
#   units::drop_units() %>%
#   mutate(Quantile = paste0(quantile * 100, "%")) %>%  # make it a label
#   transmute(
#     Quantile,
#     MAT_IRRI  = bio1.x,   MAT_Mon  = bio1.y,   MAT_GBIF  = bio1.x.x,   MAT_Arch  = bio1.y.y,
#     TMAX_IRRI = bio5.x,   TMAX_Mon = bio5.y,   TMAX_GBIF = bio5.x.x,   TMAX_Arch = bio5.y.y,
#     TWARM_IRRI= bio10.x,  TWARM_Mon= bio10.y,  TWARM_GBIF= bio10.x.x,  TWARM_Arch= bio10.y.y
#   )
# 
# gt(df_quant) %>%
#   cols_label(
#     MAT_IRRI = "IRRI",  MAT_Mon = "Mon.",  MAT_GBIF = "GBIF",  MAT_Arch = "Arch.",
#     TMAX_IRRI= "IRRI",  TMAX_Mon= "Mon.",  TMAX_GBIF= "GBIF",  TMAX_Arch= "Arch.",
#     TWARM_IRRI="IRRI",  TWARM_Mon="Mon.",  TWARM_GBIF="GBIF",  TWARM_Arch="Arch."
#   ) %>%
#   tab_spanner(label = "MAT",   columns = c(MAT_IRRI,  MAT_Mon,  MAT_GBIF,  MAT_Arch)) %>%
#   tab_spanner(label = "TMAX",  columns = c(TMAX_IRRI, TMAX_Mon, TMAX_GBIF, TMAX_Arch)) %>%
#   tab_spanner(label = "TWARM", columns = c(TWARM_IRRI,TWARM_Mon,TWARM_GBIF,TWARM_Arch)) %>%
#   fmt_number(columns = -Quantile, decimals = 1)

```

{{< pagebreak >}}

## Table S2.

```{r}
#| label: tbl-country-summaries-table
#| tbl-cap: "Projected increases in land area surpassing key temperature thresholds in each of the top 15 rice-producing countries in east Asia."
#| tbl-subcap:
#|   - "Percent land area surpassing MAT threshold"
#|   - "Percent land area surpassing TWARM threshold"
#|   - "Percent land area surpassing TMAX threshold"
#| tbl-unnumbered: true
# this is total land area??
# country_stats %>%
#   filter(variable == 'Mean annual temperature') %>%
#   select(-percent, -area) %>%
#   pivot_wider(names_from = ssp, 
#               values_from = sum) %>%
#   select(Country = name, Present = present, `SSP1-2.6`:`SSP5-8.5`) %>%
#   summarise(across(Present:`SSP5-8.5`, ~sum(.x))) %>%
#   mutate(across(everything(), ~.x / Present))
#fmt <- if (knitr::is_latex_output()) "latex" else "html"

country_stats %>%
  mutate(percent = round(percent)) %>%
  filter(variable == 'Mean annual temperature') %>%
  select(-sum, -area) %>%
  pivot_wider(names_from = ssp, values_from = percent) %>%
  select(Country = name, Present = present, `SSP1-2.6`:`SSP5-8.5`) %>%
  knitr::kable(format = 'pipe')
 # knitr::kable(format = fmt, booktabs = knitr::is_latex_output(), align = "c") |>
  #kableExtra::kable_styling(full_width = FALSE, latex_options = c("hold_position"))

country_stats %>%
  mutate(percent = round(percent)) %>%
  filter(variable == 'Max temperature of warmest quarter') %>%
  select(-sum, -area) %>%
  pivot_wider(names_from = ssp, values_from = percent) %>%
  select(Country = name, Present = present, `SSP1-2.6`:`SSP5-8.5`) %>%
  knitr::kable(format = 'pipe')
  #knitr::kable(format = fmt, booktabs = knitr::is_latex_output(), align = "c") |>
  #kableExtra::kable_styling(full_width = FALSE, latex_options = c("hold_position"))

country_stats %>%
  mutate(percent = round(percent)) %>%
  filter(variable == 'Max temperature of warmest month') %>%
  select(-sum, -area) %>%
  pivot_wider(names_from = ssp, values_from = percent) %>%
  select(Country = name, Present = present, `SSP1-2.6`:`SSP5-8.5`) %>%
  knitr::kable(format = 'pipe')
 # knitr::kable(format = fmt, booktabs = knitr::is_latex_output(), align = "c") |>
#  kableExtra::kable_styling(full_width = FALSE, latex_options = c("hold_position"))
```

{{< pagebreak >}}

## Data S1. (separate file)

`main.qmd` -- A quarto notebook containing the R code sufficient to reproduce the figures in the main analysis.

## Data S2. (separate file)

`supplementary_materials.qmd` -- A quarto notebook containing the R code sufficient to reproduce the supplementary figures and tables.

## Data S3. (separate file)

`preprocessing.qmd` -- A quarto notebook containing the R code to reproduce computationally intensive data import and preprocessing steps outlined in the Materials and Methods.
